Effective workflow in molecular dynamics simulations using Atomify --- a real-time LAMMPS visualizer

### 2018-06-20 IOPLaTeX.tex
Notes from malthe and dragly

\title{Improved workflow using Atomify - a real-time LAMMPS visualizer}

@@

\begin{abstract}
The typical workflow when running atomistic simulations includes working with several programs.
A text editor is needed to create and modify the scripts, the terminal to run the simulation, and programs like VMD or Ovito to visualize the system over time.
If physical quantities are computed, the data is often plotted using programming languages like MATLAB or Python,
where additional scripts must be used.
This is a tedious process, especially for teaching purposes and for people who are new in the field.
We here introduce Atomify; a user-friendly high performance real-time visualizer for atomistic simulations that can simulate and render more than one million atoms with excellent frame rate on modern hardware.
Atomify supports OpenMP acceleration, GPU acceleration, real-time plotting of physical quantities, and an easy-to-use code editor in one single application.
It currently uses LAMMPS as physics engine, but it can be extended to support other codes like GROMACS, NAMD or CHARMM.
Atomify is open-source software (GPL) written in C++ using the Qt framework and can be downloaded from \url{https://github.com/ovilab/atomify}.
\end{abstract}

@@

\section{Introduction}
Running molecular dynamics software has had a workflow that has not changed much over the past couple of decades.
Users of LAMMPS\cite{Plimpton1995Fast} and other codes such
as GROMACS\cite{berendsen1995gromacs}, NAMD\cite{nelson1996namd} and CHARMM\cite{brooks2009charmm} typically write
scripts in a text editor and run the scripts by passing them to the simulator on the command line.
The simulator in turn writes the results to disk and the user analyzes the results
by reading the files manually or passing them to other programs.

With the arrival of software like Rasmol\cite{sayle1995rasmol} and VMD\cite{Humphrey1996Vmd},
it became easy to both visualize the trajectories and analyze physical quantities such as the radial distribution function.
VMD came with support both for reading trajectories from file and attaching to a
running process using MDComm\cite{nelson1995mdscope} and a supported simulator such as NAMD.

Some 10 years later, an improved abstract communication layer called MDDriver\cite{delalande2009complex} was designed.
It is more general and a small effort is needed to add support for a new simulator. Another visualization tool MegaMol\cite{grottel2015megamol} was developed
only a few years ago and supports real-time visualization of GROMACS through MDDriver.
The data is transferred through the IMD protocol based on TCP/IP sockets\cite{delalande2009complex} which may introduce
overhead in data transfer compared to a shared memory solution.
Although this enables real-time visualizations, the workflow not very user-friendly since it still involves
using different programs that all need to be configured a certain way.

Analysis of simulations is often done as a postprocessing step where the trajectories and other 
per-atom quantities, such as velocity and stress, are loaded into an analysis tool written by researchers themselves.
One of today's most popular visualization tools is called OVITO\cite{Stukowski2009Visualization},
which introduced a powerful modifier pipeline where input data can be processed and manipulated as a series of scriptable building blocks.
These building blocks can be used to calculate new properties and atoms can be colored thereafter.

Scalar quantities such as temperature, pressure or number of atoms in a specific state are often stored to a
text file. Researchers then write their own plotting scripts in languages like MATLAB or Python
to plot the time evolution of these quantities.

In combination, the text editors, the simulators, the visualization software, and the programming
languages like MATLAB and Python define the workflow of a typical researcher in computational
materials science.
While this is a powerful workflow in well-established simulations,
it is tedious for prototyping new ideas.
This workflow requires a fairly large amount of context switching,
moving from the text editor to the terminal to run the simulation,
waiting for the results, and then finally being able to study them using the analysis
and visualization tools.

Here, we present Atomify, a software that integrates script editing,
visualization and analysis into one tool.
Atomify uses LAMMPS as a linked library and has shared memory access to the internals of LAMMPS,
which maximizes the datatransfer speed and reduces latency.
By having direct access to all the datastructures in LAMMPS, it is capable of not only of visualizing the simulation results,
but also show extra information about how the simulation is defined using groups and regions.

This is especially useful during development of new scripts, but can be used
to run full, publication-ready simulations with minimal simulation overhead.
The current implementation only supports LAMMPS as a simulator, but is written in a general
way so integrating other codes is a simple task.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/gui.pdf}
	\caption{%
    Overview of the Graphical User Interface (GUI) in Atomify.
    The toolbar a) is used to change between the different modes, such as
    editing, viewing and analyzing the running simulation.
    Bottom of a) contains the playback controls.
    The script editor b) supports multiple open files and provides simple syntax highlighting.
    The viewport c) shows the current simulation with atoms and bonds, simulation box and coordinate axes.
    The simulation properties d) shows the simulations details, presents
    LAMMPS objects such as groups, regions, computes, variables and fixes, and the rendering
	properties. If one of the computes, variables or fixes with plottable data is clicked, a plot window
	e) is shown with simple plot properties and export functions to MATLAB, Python or pure text file.
    }
	\label{fig:gui}
\end{figure}

In this paper we will go through a case study, an example simulation, where we
use many of the useful features Atomify provides.
We start by discussing the GUI in section \ref{sec:features} where the main features are summarized.
In section \ref{sec:casestudy}, a case study is introduced where we go through a somewhat complicated simulation
that uses regions and groups to perform different integration rules on different atoms.

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.5\textwidth]{figures/rightbar.pdf}
	\caption{
		The \textit{Simulation} panel is a powerful feature in Atomify where relevant
		properties from LAMMPS is presented real-time.
		In a), a list of all groups defined is shown with the current atom count.
		b) shows all defined regions. Each group and region can be hovered so that atoms
		are colored after what group / region they are in. If an atom is in multiple
		groups or regions, they will get the color of the latest one defined.
		The color boxes can be clicked which will either always highlight the atoms in a certain group or region,
		or if clicked again, hide all of them.
		In c) and d), a list of all variables and computes is shown. If a variable or compute
		produces a scalar, the current value is shown as can be seen with all the variables.
		By clicking the name, a plot will show up showing the time evolution of that quantity.
		If a per-atom variable or compute is hovered, all atoms will be colored according to their
		current scalar value and a histogram over these values is shown if clicked.
		The result of hovering atoms defined by the groups in our case study is shown in figure \ref{fig:cylinder_simulation}.
    }
	\label{fig:rightbar}
\end{figure}

@@

Atomify was designed with one main goal: improve the workflow.
Focusing on minimizing the number of context switches and mouse clicks reduces
the distance between creativity and simulation results.
It has a script editor, visualization window and a simulation summary panel so all relevant parts
of a simulation can be on the screen at the same time.

Any changes to the script like defining regions or groups,
changing temperatures, size of simulation box or inserting atoms can
quickly be verified since the visualization shows the real-time state of the simulator.
Rerunning the current script can be done by pressing \keys{Ctrl+R} (\keys{\cmd+R} on macOS).

How the pressure, radial distribution function or some other
complicated variable evolves over time is easily found
since these can be plotted real-time with a simple click in the list of variables.
Bonds between atoms in a molecule are automatically extracted from LAMMPS, or can
be specified as a distance threshold between atom types.
Regions and groups can be highlighted or hidden, which is useful to focus on some parts of a system.
Rendering of periodic copies helps studying features near a periodic boundary.

@@

Although LAMMPS is mostly used to study molecular dynamics, it can also be used to
study monte carlo simulations\cite{frenkel2001understanding} and granular
materials\cite{brilliantov1996model, silbert2001granular, zhang2005jamming},
both of which work very well in Atomify since they are particle based simulations.

Atomify supports rendering of spheres and cylinders representing atoms and bonds.
It is done using the powerful Qt3D engine with custom visualization techniques such as
billboard raycasting\cite{gumhold2003splatting, sigg2006gpu, tarini2006ambient} and screen space ambient occlusion\cite{bavoil2008screen}
for improved dynamic depth perception. This allows a large amount of atoms and molecules being rendered with excellent framerate.
The liming factor is the simulation itself, since force calculation is a heavy task.

This turns out to improve the workflow substantially.
In addition, Atomify comes with multiple examples,
many of which have been made by members of the LAMMPS community.

\section{\label{sec:casestudy}Case study: flow through a narrow nanotube}
Atomify is best demonstrated with a case study where many of
the useful features are used to obtain a productive workflow during script development.
We have chosen to create a system with fluid flow inside a tight nanochannel.
This involves creating a somewhat complicated initial geometry with multiple regions
where we apply different rules such as a dissipative thermostat, constant force mimicking
applied pressure gradient, and frozen atoms to avoid net momentum of the solid.
We will create two different atom types so we can have a gas phase and a solid phase at the same temperature.

Fluid flow is often characterized with its permeability - the inverse flow resistance of a material.
The permeability can be measured using Darcy's law when applying a pressure gradient $\nabla P$.
In Darcy's law, the permeability is assumed to only depend on the material and its geometry.
However, this turns out not being the case for dilute
gases where the measured permeability changes as a function of pressure\cite{klinkenberg1941permeability}.
This is called the Klinkenberg effect.

The discrepancy is explained by the fact that at low pressures, the fluid velocity
at the boundary is non-zero, i.e. we have a non-slip boundary condition.
Darcy's law was found as an empirical law using water and sand.
A non-slip boundary condition is a good approximation for liquids in larger pores,
but if the channel size $R$ is of the same order,
or larger, as the mean free path $\lambda$ of the fluid, it is no longer true.

The ratio between the mean free path and the channel size is quantified through
the Knudsen number, $\text{Kn} = \lambda / R$, and can be used to apply the Klinkenberg correction\cite{klinkenberg1941permeability}
to correctly describe the \textit{effective permeability} as a function of pressure in the high Knudsen number limit.

@@

In this case study, we will use Atomify to measure the velocity profile of a
gas inside a nanotube with varying density to observe
the slip velocity effect at low densities.
We will use the Lennard Jones potential with two different atom types.

We create a nanotube inside a solid with  $(60\times30\times30)$ FCC unit cells and reduced density $\rho^* = 0.975$.
This density was chosen so that the two atom types with the Lennard Jones coefficients specified below will coexist in different phases.
The fluid in the nanotube, or cylinder, will flow due to an induced pressure gradient, see figure \ref{fig:cylinder_simulation}.
Here we apply a pressure gradient $\nabla P$ in the $x$-direction on the gas atoms using a constant acceleration $a$ calculated as
\[
    a = \frac{\nabla P}{\rho_m},
\]
where $\rho_m$ is the mass density of the gas atoms and $\nabla P$ is the desired pressure gradient.
This represents a pressure difference $\Delta P$ over a small volume element $\Delta x$ on a cross-sectional area $\Delta y\Delta z$.

Applying an acceleration means adding energy to the system which in turn will increase the temperature and eventually melt the system or make it move.
To avoid this, we couple the solid to an external heat bath so the added energy can be transferred out from the system through a dissipative thermostat.

Even though we only apply a pressure gradient on the gas atoms, the induced momentum will be transferred
to the solid atoms which eventually will start moving in the positive $x$-direction.
We want the solid to be at rest since it is only the relative velocity between the gas and the solid that defines
the flow. We therefore freeze the outmost atoms (colored cyan in Fig \ref{fig:cylinder_simulation}).
To obtain this system, we divide the atoms into multiple cylindrical regions, which are integrated differently:

\begin{itemize}
	\item $r \leq 12$: gas atoms with applied pressure gradient
	\item $18 < r \leq 23$: dissipative Langevin thermostat
	\item $r \leq 23$: regular time integration
	\item $23 < r$: frozen.
\end{itemize}
The different radii are specified in reduced Lennard Jones units.

@@

We will modify this script bit by bit.
Here, we have a Lennard Jones solid with reduced density
$\rho^* = 0.975$, reduced masses $m_1^* = 4.0$, $m_2^* = 0.5$ with $(60\times30\times30)$
FCC unit cells and a cutoff of $2.5$. 
The force field parameters are chosen so that atoms of type 1
are in the solid phase while type 2 atoms are gas at reduced temperature $T^*=0.5$.
The two atom types are weakly coupled (small $\epsilon_{ij}$) so the gas atoms won't
stick too easily to the surface since this would prevent flow.

To make sure LAMMPS executes our modifiction before visualization is started,
we need to make sure that following additional commands happen \textit{before} the \code{run 1000} command.
If not, LAMMPS will run 1000 timesteps before getting to our changes.

To define the inner cylinder, we first specify the radius and the center before using the region command
\begin{lstlisting}
variable R equal 12
variable c equal $(ly*0.5)
region cylinder1 cylinder x $c $c $R EDGE EDGE
set region cylinder1 type 2
\end{lstlisting}
The region command will create a cylinder $C_1$ in the $x$-direction, with $(y,z)$-coordinates at the 
system center \code{\$c} with radius \code{\$R} using the full length of the system in the $x$-direction specified with \code{EDGE}.
We also change the atoms inside the region to type 2 since are the gas atoms.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{figures/initial_configuration.pdf}
	\caption{
		Several snapshots of the process of creating the simulation in our case study.
		This shows how one typically generate a geometry step by step where the immediate
		feedback is very useful to quickly discover mistakes.
		The wrongly placed cylinder is shown in a) whereas it is placed correctly in b).
		c) shows how the system looks after deleting atoms to obtain $\rho_m = 0.05$. The next cylinder $C_2$ is shown in d),
		and all regions (placed in groups) are shown in e). Here we have uniquely defined the purple shell which should have a dissipative thermostat.
		In f), the list of all groups defined in this system is shown with colors matching those of e).
		Although one group is colored green, we don't see any green atoms (\code{cylinder1})
		since all of them also are in the groups \code{cylinder2}, \code{cylinder3}, \code{nve} and \code{gas},
		which all override the color. The color of an atom is determined by the last defined group it is member of.
		All figures are rendered with Atomify.
    }
	\label{fig:initial_configuration}
\end{figure}

When we hover the region, we immediately see that the cylinder is in fact not placed at the center of the system, as shown in figure \ref{fig:initial_configuration}a.
The reason is that LAMMPS operate with two different units, either the length unit or the lattice length unit. This is explained in the documentation, but is a common mistake.
Discovering mistakes like this is easy using Atomify.

The problem is that LAMMPS expects the values to be in \textit{lattice} units, i.e. multiples of the lattice constant.
Since we used \code{\$(0.5*lx)}, which gives the system length in actual lenght units, we need to specify \code{units box} in the region command.
We then rerun the simulation with \keys{Ctrl+R} which gives the expected result in figure \ref{fig:initial_configuration}b.

Now that the region looks right, we want to delete some atoms to control the density.
The volume of the inner cylinder is $\pi R^2 L$ which can be calculated in LAMMPS with \code{variable V equal PI*v\_R*v\_R*lx}.
The command \code{delete\_atoms} can be used: \code{delete\_atoms porosity region-ID fraction seed},
where we need to specify the fraction of the existing atoms in that region we want to delete, and a random seed which can be any positive number.
Since we want to specify a certain density, we need to figure out how many atoms we
should delete. This depends on the current number of atoms inside the region.

@@

This shows a powerful workflow with Atomify where changes quickly can be tested with immediate feedback.
Each time we add or change a command, we just rerun with \keys{Ctrl+R} to see how it looks like.

The next step now is to create a larger cylinder $C_2$, the one containing the orange atoms in
figure \ref{fig:cylinder_simulation}, where regular time integration will take place.
This cylinder should have a slightly bigger radius and can be created with another region command:
\begin{lstlisting}
region cylinder2 cylinder x $c $c $(v_R+6) EDGE EDGE units box
\end{lstlisting}
Again press \keys{Ctrl+R} to see the expected result in figure \ref{fig:initial_configuration}d.

Now we have the inner cylinder with the gas atoms defined as a region and the next cylinder shell (including the gas atoms) defined as another region.
We now want to create another shell, the one with the dissipative thermostat.
All the atoms with the dissipative thermostat, the inner shell and the gas will be integrated with the \code{fix nve} integrator.
We therefore create a larger cylinder $C_3$ with $r=21$ which will be the one defining the group for \code{fix nve}.
With all three regions defined, we also create groups which will be used in fixes
\begin{lstlisting}
region cylinder3 cylinder x $c $c $(v_R+9) EDGE EDGE units box
group cylinder1 region cylinder1
group cylinder2 region cylinder2
group cylinder3 region cylinder3
group nve region cylinder3
\end{lstlisting}
where we also have created the group \code{nve} for clarity.
Note that cylinder 3 contains all atoms in cylinder 2 which also contains all atoms in cylinder 1.

@@

We here also created the group \code{frozen}, which contains only the outmost atoms,
and the group \code{gas} which are all atoms of type 2.
The frozen atoms will not move at all and make sure that the system does not start 
moving due to momentum transfer from the flowing gas atoms.

@@

The syntax for fixes is \code{fix ID groupID style args}, so the above fixes have the same name as both the groups and the fix style.
\code{fix langevin} takes at least 4 arguments \textit{Tstart, Tstop, Tdamp, seed}, i.e.
the thermostat temperature in the beginning of a simulation, final thermostat temperature at the end, a damping parameter which controls how strongly the thermal bath interacts with the system.
It also needs a random seed due to the random nature of the Langevin thermostat as discussed in \cite{schneider1978molecular}.
Finally, we have added another compute that measures the displacement per atom which we will use to see if atoms are placed stable in the crystal lattice.
In Atomify, we can hover this compute in the list of computes which will color each atom according to their displacement since $t=0$.
This works for any \textit{per-atom} compute or variable.

>>>>

\title{Effective workflow in molecular dynamics simulations using Atomify --- a real-time LAMMPS visualizer}

@@

\begin{abstract}
A typical workflow when setting up and running atomistic simulations combines the use of several tools:
a text editor to create and modify scripts; the terminal to run the simulation; and programs like VMD or Ovito to visualize the dynamics of the system. In addition, to study the behavior of physical quantities, relevant data must be computed, stored, and subsequently plotted. This is a cumbersome process, which limits the use of atomic simulation tools for teaching and increases the barrier of adaptation for newcomers. 
Here, we introduce Atomify to streamline the atomic simulation workflow. It is a user-friendly high-performance real-time visualizer for atomistic simulations that can simulate and render more than one million atoms with excellent frame rate on modern hardware.
Atomify supports OpenMP acceleration, GPU acceleration, real-time plotting of physical quantities, and an easy-to-use code editor in one single application.
It currently uses LAMMPS as physics engine, but it can be extended to support other underlying codes like GROMACS, NAMD or CHARMM.
Atomify is open-source software (GPL) written in C++ using the Qt framework and can be downloaded from \url{https://github.com/ovilab/atomify}.
\end{abstract}

@@

\section{Introduction}
Atomic scale simulations using molecular dynamics simulations have become a standard tool in scientific research, development and education\cite{frenkel2001understanding,StatisticalAndThermalPhysics-Gould-2006}.
However, the workflow to set up and run molecular dynamics simulations has remained mostly unchanged over the past decades.
Users of LAMMPS\cite{Plimpton1995Fast}, GROMACS\cite{berendsen1995gromacs}, NAMD\cite{nelson1996namd} and CHARMM\cite{brooks2009charmm} typically write
scripts in a text editor, run the scripts by passing them to the simulator on the command line. 
The simulator in turn writes the results to disk, and the user then analyzes the results
by reading the files manually or passing them to other programs.

With the introduction of software like Rasmol\cite{sayle1995rasmol} and VMD\cite{Humphrey1996Vmd},
it became easy to both visualize simulation trajectories and analyze calculated quantities such as the radial distribution function within the same tool. 
VMD came with support both for reading trajectories from file and attaching to a
running process using MDComm\cite{nelson1995mdscope} and a supported simulator such as NAMD.

Some 10 years later, an improved abstract communication layer called MDDriver\cite{delalande2009complex} was designed.
Compared to MDComm it was more general and required only a small effort to add support for a new simulator. Another visualization tool MegaMol\cite{grottel2015megamol} was developed more recently and supported real-time visualization of GROMACS through MDDriver.
However, data was transferred through the IMD protocol based on TCP/IP sockets\cite{delalande2009complex}, and this may have introduced
overhead in the data transfer compared to a shared memory solution.
Although this enabled real-time visualizations, the workflow was not very user-friendly since it still involved
the use of different programs that all needed to be configured a particular way.

Analysis of simulations is often done as a postprocessing step where the trajectories and other 
per-atom quantities, such as velocity and stress, are loaded into an analysis tool often written by researchers themselves.
One of today's most popular visualization tools for atomistic simulations is OVITO\cite{Stukowski2009Visualization},
which introduced a powerful modifier pipeline where input data can be processed and manipulated as a series of scriptable building blocks.
These building blocks can be used to calculate new properties, and atoms can be colored accordingly.

In addition, aggregate scalar quantities such as temperature, pressure or number of atoms in a specific state are often stored to a
text file. Researchers then write their own plotting scripts in languages like MATLAB or Python
to plot the time evolution of these quantities.

In combination, the text editors, the simulators, the visualization software, and MATLAB or Python scripts define the workflow of a typical researcher in computational
materials science.
While this is a powerful workflow in well-established simulation cases,
it is tedious for prototyping new ideas.
This workflow requires a fairly large amount of context switching,
moving from the text editor to the terminal to run the simulation,
waiting for the results, and then finally being able to study them using the analysis
and visualization tools.

Here, we present Atomify, a software that integrates script editing,
visualization and analysis into one tool.
Atomify uses LAMMPS as a linked library and has shared memory access to the internals of LAMMPS,
which maximizes the data transfer speed and reduces latency.
By having direct access to all the data structures in LAMMPS, Atomify is capable of not only visualizing the simulation results,
but also show additional information about how the simulation is defined using groups and regions.

This is especially useful during development of new scripts, but can also be used
to run full, publication-ready simulations with minimal simulation overhead.
The current implementation only supports LAMMPS as a simulator, but it is written in a general
way so integrating other codes is a simple task.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/gui.pdf}
	\caption{%
	Overview of the Graphical User Interface (GUI) in Atomify.
	The toolbar a) is used to change between the different modes, such as
	editing, viewing and analyzing the running simulation.
	The bottom of a) contains the playback controls.
	The script editor b) supports multiple open files and provides simple syntax highlighting.
	The viewport c) shows the current simulation with atoms and bonds, simulation box and coordinate axes.
	The simulation properties d) shows the simulations details, presents
	LAMMPS objects such as groups, regions, computes, variables and fixes, and the rendering
	properties. If one of the computes, variables or fixes with plottable data is clicked, a plot window
	e) is shown with simple plot properties and export functions to MATLAB, Python or a text file.
	}
	\label{fig:gui}
\end{figure}

In this paper we will 
introduce and demonstrate many of the useful features that Atomify provides through a case study. 
We start by introducing the GUI in section \ref{sec:features}, which also summarizes the main features of Atomify.
In section \ref{sec:casestudy}, the case study is introduced. Here, we provide a detailed walk-through of setting up, running and analyzing a non-trivial simluation that uses regions and groups to perform different integration rules on different atoms.

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.5\textwidth]{figures/rightbar.pdf}
	\caption{
		The \textit{Simulation} panel is a powerful feature in Atomify where relevant
		properties from LAMMPS are presented real-time.
		In a), a list of all defined groups is shown with the current atom count.
		b) shows all defined regions. Each group and region can be hovered with the mouse so that atoms
		are colored after what group/region they are in. If an atom is in multiple
		groups or regions, they will get the color of the latest one defined.
		The color boxes can be clicked to either always highlight the atoms in a certain group or region,
		or if clicked again, hide all of them.
		In c) and d), a list of all variables and computes is shown. If a variable or a compute
		produce a scalar, the current value is shown as can be seen with all the variables.
		By clicking the name, a plot will show up with the time evolution of that quantity.
		If a per-atom variable or compute is hovered, all atoms will be colored according to their
		current scalar value and a histogram over these values is shown if clicked.
		The result of hovering atoms defined by the groups in our case study is shown in figure \ref{fig:cylinder_simulation}.
	}
	\label{fig:rightbar}
\end{figure}

@@

Atomify was designed with one goal at mind: to improve the workflow.
By focusing on minimizing the number of context switches and mouse clicks, 
the distance between creativity and simulation results is reduced. 
Atomify has a script editor, a visualization window and a simulation summary panel enabling all relevant parts
of a simulation to be on the screen at the same time.

The effect of any changes to the script, such as defining regions or groups,
changing temperatures or the size of the simulation box, or inserting atoms, can
quickly be checked since the visualization shows the real-time state of the simulator.
Rerunning the current script can be done by pressing \keys{Ctrl+R} (\keys{\cmd+R} on macOS).

The temporal evolution of the pressure, the radial distribution function or any other
variable can be plotted real-time with a simple click in the list of variables.
Bonds between atoms in a molecule are automatically extracted from LAMMPS, or can
be specified as a distance threshold between atom types.
Regions and groups can be highlighted or hidden, which is useful to focus on some parts of a system.
Rendering of periodic copies helps studying features near a periodic boundary.

@@

Although LAMMPS is mostly used to study molecular dynamics, it can also be used to
model Monte Carlo simulations\cite{frenkel2001understanding} and granular
materials\cite{brilliantov1996model, silbert2001granular, zhang2005jamming},
both of which work in Atomify since they are particle based simulations.

Atomify supports rendering of spheres and cylinders to represent atoms and bonds.
It is done using Qt3D with custom visualization techniques such as
billboard raycasting\cite{gumhold2003splatting, sigg2006gpu, tarini2006ambient} and screen space ambient occlusion\cite{bavoil2008screen}
for improved dynamic depth perception. This allows a large amount of atoms and molecules to be rendered with excellent framerate.
The liming factor is the simulation itself, since force calculation is a computationally intensive task.

In addition, Atomify comes with substantial database of examples,
many of which have been made by members of the LAMMPS community.

\section{\label{sec:casestudy}Case study: flow through a narrow nanotube}
Atomify is best demonstrated through a case study that shows how 
various features can be used in a productive workflow from script development to analysis.
We have chosen to address fluid flow inside a narrow, cylindrical nanochannel, since this is a scenario that demonstrates many of the features of Atomify and its integrated workflow: The simulation involves creating an initial geometry with multiple regions 
where we apply different rules such as a dissipative thermostat, a constant force mimicking
an applied pressure gradient, and frozen atoms to avoid net momentum of the solid.
We introduce two different atom types in order to have a gas phase and a solid phase at the same temperature.

Fluid flow in porous structures is often characterized by its permeability, $k$, which describes the fluid flow conductivity of a fluid in a specific geometry. Darcy's law relates the average flow velocity, $U$, to the applied pressure gradient $\nabla P$: $U = (k/\mu)(-\nabla P)$, where $\mu$ is the viscosity of the fluid. In Darcy's law, the permeability is assumed to only depend on the material and its geometry. 
\todo[inline]{U is the flux here, not the average velocity?}
However, Darcy's law must be modified for dilute
gases where the measured permeability also is found to also be a function of pressure.
This is called the Klinkenberg effect\cite{klinkenberg1941permeability}.

The discrepancy is due to finite slip of the dilute gas along the pore walls. For a macroscopic fluid, the velocity of the fluid is usually assumed to be zero at the pore walls. This is called non-slip boundary conditions and is a good approximation for dense liquids in macroscopic pores. However, if the typical channel size $R$ in the pores is of the same order as the mean free path $\lambda$ of the fluid, non-slip boundary conditions are no longer a good approximation to the actual flow profile. The ratio between the mean free path and the channel size is quantified through
the Knudsen number, $\text{Kn} = \lambda / R$ is used in the Klinkenberg correction\cite{klinkenberg1941permeability} for the \textit{effective permeability} as a function of pressure in the limit of high Knudsen numbers. Such a system is well suited for a molecular dynamics study because channles with small values of $R$, and hence high Knudsen numbers, may be studied in molecular systems.

@@

\subsection{Physical description of the simulation system}
In this case study, we will use Atomify to measure the velocity profile of a
gas inside a nanotube with varying density to directly observe
the slip velocity at low densities.
We will use the Lennard-Jones potential with two different atom types.

We create a nanotube inside a FCC solid with $(60\times30\times30)$ unit cells and reduced density $\rho^* = 0.975$.
The density was chosen to ensure that the two atom types with the Lennard-Jones coefficients specified below will coexist in different phases. We will apply a pressure gradient to the fluid inside the nanotube to induce fluid flow, see figure \ref{fig:cylinder_simulation}.
We apply a pressure gradient $\nabla P$ in the $x$-direction on the gas atoms using an added force, which corresponds to an acceleration $a$ added to each atoms. The acceleration $a$ calculated as
\[
	a = \frac{\nabla P}{\rho_m},
\]
where $\rho_m$ is the mass density of the gas atoms, produces the desired pressure gradient $\nabla P$.
This represents a pressure difference $\Delta P$ over a small volume element $\Delta x$ on a cross-sectional area $\Delta y\Delta z$.

By adding an external force to atoms in the system, energy is added to the system, which in turn will increase the temperature and eventually melt the system or make it move.
To avoid this, we couple the solid to an external heat bath so the added energy can be transferred out from the system through a dissipative thermostat.

Even though we only apply a pressure gradient on the gas atoms, the induced momentum will also be transferred
to the solid atoms, which eventually will start moving in the positive $x$-direction.
We want the solid to be at rest, since only the relative velocity between the gas and the solid defines
the flow. We will therefore freeze the outermost atoms (colored cyan in Fig \ref{fig:cylinder_simulation}).

To obtain this system, we divide the atoms into multiple cylindrical regions, which are integrated differently:

\begin{itemize}
	\item $r \leq 12$: gas atoms with applied pressure gradient
	\item $18 < r \leq 21$: dissipative Langevin thermostat
	\item $21 < r$: frozen.
\end{itemize}

The different radii are specified in reduced Lennard-Jones units.

In the following we will provide a detailed description of how to set up, run and analyze this system using Atomify.

@@

We will modify and adapt this script to bit by bit to correspond to the system we want to simulate.
Here, we have a Lennard-Jones solid with reduced density
$\rho^* = 0.975$, reduced masses $m_1^* = 4.0$, $m_2^* = 0.5$ with $(60\times30\times30)$
FCC unit cells and a cutoff of $2.5$. 
The force field parameters are chosen so that atoms of type 1
are in the solid phase while type 2 atoms are gas at reduced temperature $T^*=0.5$.
The two atom types are weakly coupled (small $\epsilon_{ij}$) so the gas atoms won't
stick too easily to the surface since this would prevent flow.

To make sure LAMMPS executes our modification before visualization is started,
we need to make sure that following additional commands happen \textit{before} the \code{run 1000} command.
If not, LAMMPS will run 1000 timesteps before getting to our changes.

To define the inner cylinder, we first specify the radius and the center before using the region command
\begin{lstlisting}
variable R equal 12
variable c equal $(ly*0.5)
region cylinder1 cylinder x $c $c $R EDGE EDGE
set region cylinder1 type 2
\end{lstlisting}
The region command will create a cylinder $C_1$ in the $x$-direction, with $(y,z)$-coordinates at the 
system center \code{\$c} with radius \code{\$R} using the full length of the system in the $x$-direction specified with \code{EDGE}.
We also change the atoms inside the region to type 2 since these are the gas atoms.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{figures/initial_configuration.pdf}
	\caption{
		Several snapshots of the process of creating the simulation in our case study.
		This shows how one typically generates a geometry step by step where the immediate
		feedback is very useful to quickly discover mistakes.
		The wrongly placed cylinder is shown in a) whereas it is placed correctly in b).
		c) shows how the system looks after deleting atoms to obtain $\rho_m = 0.05$. The next cylinder $C_2$ is shown in d),
		and all regions (placed in groups) are shown in e). Here we have uniquely defined the purple shell which should have a dissipative thermostat.
		In f), the list of all groups defined in this system is shown with colors matching those of e).
		Although one group is colored green, we don't see any green atoms (\code{cylinder1})
		since all of them also are in the groups \code{cylinder2}, \code{cylinder3}, \code{nve} and \code{gas},
		which all override the color. The color of an atom is determined by the last defined group it is a member of.
		All figures are rendered with Atomify.
	}
	\label{fig:initial_configuration}
\end{figure}

When we hover the region, we immediately see that the cylinder is not placed at the center of the system, as shown in figure \ref{fig:initial_configuration}a.
The reason is that LAMMPS operates with two different units, either the length unit or the lattice length unit. This is explained in the documentation, but is a common mistake.
Discovering mistakes like this is easy using Atomify.

Since we used \code{\$(0.5*lx)}, which gives the system length in actual lenght units, we need to specify \code{units box} in the region command:
\begin{lstlisting}
	region cylinder1 cylinder x $c $c $R EDGE EDGE units box
\end{lstlisting}
We then rerun the simulation with \keys{Ctrl+R} and see that we get the wanted result in figure \ref{fig:initial_configuration}b.

Now that the region looks right, we want to delete some atoms to control the density.
The volume of the inner cylinder is $\pi R^2 L$ which can be calculated in LAMMPS with \code{variable V equal PI*v\_R*v\_R*lx}.
The command \code{delete\_atoms} can be used: \code{delete\_atoms porosity region-ID fraction seed},
where we need to specify what fraction of the existing atoms in the region that we want to delete, and a random seed which can be any positive number.
Since we want to specify a specific density, we need to figure out how many atoms we
should delete. This depends on the current number of atoms inside the region.

@@

This shows a powerful workflow with Atomify where changes quickly can be tested with immediate feedback.
Each time we add or change a command, we just rerun with \keys{Ctrl+R} to see what the results look like.

The next step is to create a larger cylinder $C_2$ containing the orange atoms in
figure \ref{fig:cylinder_simulation}, where regular time integration will take place.
This cylinder should have a slightly bigger radius and can be created with another region command:
%
\begin{lstlisting}
region cylinder2 cylinder x $c $c $(v_R+6) EDGE EDGE units box
\end{lstlisting}
%
Again press \keys{Ctrl+R} to see the expected result in figure \ref{fig:initial_configuration}d.
Now we have the inner cylinder with the gas atoms defined as a region and the next cylinder shell (including the gas atoms) defined as another region.
We now want to create another shell, the one with the dissipative thermostat.
All the atoms with the dissipative thermostat, the inner shell and the gas will be integrated with the \code{fix nve} integrator.
We therefore create a larger cylinder $C_3$ with $r=21$, which will be the one defining the group for \code{fix nve}.
With all three regions defined, we also create groups which will be used in fixes
\begin{lstlisting}
region cylinder3 cylinder x $c $c $(v_R+9) EDGE EDGE units box
group cylinder1 region cylinder1
group cylinder2 region cylinder2
group cylinder3 region cylinder3
group nve region cylinder3
\end{lstlisting}
where we also have created the group \code{nve} for clarity.
Note that cylinder 3 contains all atoms in cylinder 2 which also contains all atoms in cylinder 1.

@@

We here also created the group \code{frozen}, which contains only the outermost atoms,
and the group \code{gas}, which are all atoms of type 2.
The frozen atoms will not move at all and make sure that the system does not start 
moving due to momentum transfer from the flowing gas atoms.

@@

The syntax for fixes is \code{fix ID groupID style args}, so the above fixes have the same name as both the groups and the fix style.
\code{fix langevin} takes at least 4 arguments \textit{Tstart, Tstop, Tdamp, seed}, i.e.
the thermostat temperature in the beginning of a simulation, final thermostat temperature at the end, and a damping parameter, which controls how strongly the thermal bath interacts with the system.
It also needs a random seed due to the random nature of the Langevin thermostat as discussed in \cite{schneider1978molecular}.
Finally, we have added another compute that measures the displacement per atom which we will use to see if atoms remain stable in the crystal lattice.
In Atomify, we can hover this compute in the list of computes to color each atom according to their displacement since $t=0$.
